A memory frontier for complex synapses

### 2013-11-06 NIPS/SynapseNIPS.tex
typos
final NIPS paper

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{A memory frontier for complex synapses}
%
% Author List:
%
\author{Subhaneil Lahiri and Surya Ganguli\\
Applied Physics Department, Stanford University, Stanford CA\\
\emaillink{sulahiri@stanford.edu}, \emaillink{sganguli@stanford.edu}
%
}

@@

Theoretical studies of complex, metaplastic synapses, have focused on analyzing the memory performance of a limited number of very specific, molecular dynamical systems, characterized by a number of internal states in which potentiation and depression each induce a specific set of allowable transitions between states (\eg see \autoref{fig:models} below).
While these models can vastly outperform simple binary synaptic switches, these analyses leave open several deep and important questions.
For example, how does the structure of a synaptic dynamical system determine its memory performance?  What are the fundamental limits of memory performance over the space of all possible synaptic dynamical systems?
What is the structural organization of synaptic dynamical systems that achieve these limits?
Moreover, from an experimental perspective, it is unlikely that all synapses can be described by a single canonical synaptic model;  just like the case of neurons, there is an incredible diversity of molecular networks underlying synapses both across species and across brain regions within a single organism \cite{Emes2012synapserev}.
In order to elucidate the functional contribution of this diverse molecular complexity to learning and memory, it is essential to move beyond the analysis of specific models and instead develop a general theory of learning and memory for complex synapses.
Moreover, such a general theory of complex synapses could aid in development of novel artificial memory storage devices.  % Sutskever, memory for learning?

@@

A simple mean memory curve can be derived as follows.
All of the preceding plasticity events, prior to $t=0$, will put the population of synapses in its steady-state distribution, $\eq$.
The memory we are tracking at $t=0$ will change the internal state distribution to $\eq\M\pot$ in those synapses that are potentiated and $\eq\M\dep$ in those synapses that are depressed.
As the potentiating/depressing nature of the subsequent memories is independent of $\wvi$, we can average over all sequences, resulting in the evolution of the probability distribution:
%
\begin{equation}\label{eq:evol}
  \diff{\pr(t)}{t} = r\pr(t)\frg,
  \qquad \text{where} \quad
  \frg = f\pot\M\pot + f\dep\M\dep - \I.
\end{equation}
%
Here $\frg$ is a continuous time transition matrix that models the process of forgetting the memory stored at time $t=0$ due to random candidate potentiation/depression events occurring at each synapse due to the storage of subsequent memories.
Its stationary distribution is $\eq$.

@@

Much of the previous work on these types of complex synaptic models has focused on understanding the memory curves of specific models, or choices of $\M\potdep$.
Two examples of these models are shown in \autoref{fig:models}. We see that they have different memory properties.
The serial model performs relatively well at one particular timescale, but it performs poorly at other times.
The cascade model does not perform quite as well at that time, but it maintains it performance over a wider range of timescales.

@@

We now give an upper bound on the initial SNR,
%
\begin{equation}\label{eq:init}
  \snr(0) = \sqrt{N}\prn{2 f\pot f\dep} \eq \prn{\M\pot-\M\dep} \w,
\end{equation}
%
over \emph{all} possible models and also find the model that saturates this bound.
A useful quantity is the equilibrium probability flux between two disjoint sets of states, $\CA$ and $\CB$:
%
\begin{equation}\label{eq:flux}
  \F_{\CA\CB} = \sum_{i\in\CA} \sum_{j\in\CB} r \eq_i \frg_{ij}.
\end{equation}
%
The initial SNR is closely related to the flux from the states with $\w_i=-1$ to those with $\w_j=+1$ (see supplementary material):
%
\begin{equation}\label{eq:initflux}
  \snr(0) \leq \frac{4\sqrt{N}\F_{-+}}{r}\,.
\end{equation}
%
This inequality becomes an equality if potentiation never decreases the synaptic weight and depression never increases it, which should be a property of any sensible model.

@@

Now consider the area under the memory curve:
%
\begin{equation}\label{eq:area}
  A = \int_0^\infty\!\!\dr t \, \snr(t).
\end{equation}
%
We will find an upper bound on this quantity as well as the model that saturates it.

@@

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \autoref{fig:perts}, see supplementary material for details.
For example, in \autoref{fig:perts}(\ref{fig:tri_right}), for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\smash{\eta^+_i > \eta^+_j}$), we have proven that increasing $\smash{\M\pot_{ij}}$ and decreasing $\smash{\M\dep_{ij}}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \autoref{fig:env}(\ref{fig:sticky}),%\autoref{fig:max_area},
with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some perturbation to further increase the area.

@@

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$ (see supplementary material):
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn[{k - \sum_j j\eq_j}]$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \autoref{fig:env}(\ref{fig:sticky}).%\autoref{fig:max_area}.
This makes it very difficult to get out of the end states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

@@

This leaves the question of whether this bound is achievable.
At any time, can we find a model whose memory curve touched the envelope?
The red curves in \autoref{fig:env}(\ref{fig:envplot}) show the closest we have come to the envelope with actual models, by repeated numerical optimization of $\SNR(t_0)$ over $\M\potdep$ with random initialization and by hand designed models.

@@

%Another important consideration is the tolerance of the synapses to biological noise, which could be investigated by imposing lower limits on the level of stochasticity of the models.
Finally, it would be interesting to design physiological experiments in order to perform optimal systems identification of potential Markovian dynamical systems hiding within biological synapses,
given measurements of pre and post-synaptic spike trains along with changes in post-synaptic potentials.
Then given our theory, we could match this measured synaptic model to optimal models to infer which timescales the synapse operates on.

>>>>

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{A memory frontier for complex synapses}
%
% Author List:
%
\author{Subhaneil Lahiri and Surya Ganguli\\
Department of Applied Physics, Stanford University, Stanford CA\\
\emaillink{sulahiri@stanford.edu}, \emaillink{sganguli@stanford.edu}
%
}

@@

Theoretical studies of complex, metaplastic synapses have focused on analyzing the memory performance of a limited number of very specific molecular dynamical systems, characterized by a number of internal states in which potentiation and depression each induce a specific set of allowable transitions between states (\eg see \autoref{fig:models} below).
While these models can vastly outperform simple binary synaptic switches, these analyses leave open several deep and important questions.
For example, how does the structure of a synaptic dynamical system determine its memory performance?  What are the fundamental limits of memory performance over the space of all possible synaptic dynamical systems?
What is the structural organization of synaptic dynamical systems that achieve these limits?
Moreover, from an experimental perspective, it is unlikely that all synapses can be described by a single canonical synaptic model;  just like the case of neurons, there is an incredible diversity of molecular networks underlying synapses both across species and across brain regions within a single organism \cite{Emes2012synapserev}.
In order to elucidate the functional contribution of this diverse molecular complexity to learning and memory, it is essential to move beyond the analysis of specific models and instead develop a general theory of learning and memory for complex synapses.
Moreover, such a general theory of complex synapses could aid in development of novel artificial memory storage devices.  % Sutskever, memory for learning?

@@

A simple average memory curve can be derived as follows.
All of the preceding plasticity events, prior to $t=0$, will put the population of synapses in its steady-state distribution, $\eq$.
The memory we are tracking at $t=0$ will change the internal state distribution to $\eq\M\pot$ (or $\eq\M\dep$) in those synapses that experience a candidate potentiation (or depression) event.
As the potentiating/depressing nature of the subsequent memories is independent of $\wvi$, we can average over all sequences, resulting in the evolution of the probability distribution:
%
\begin{equation}\label{eq:evol}
  \diff{\pr(t)}{t} = r\pr(t)\frg,
  \qquad \text{where} \quad
  \frg = f\pot\M\pot + f\dep\M\dep - \I.
\end{equation}
%
Here $\frg$ is a continuous time transition matrix that models the process of forgetting the memory stored at time $t=0$ due to random candidate potentiation/depression events occurring at each synapse due to the storage of subsequent memories.
Its stationary distribution is $\eq$.

@@

Much of the previous work on these types of complex synaptic models has focused on understanding the memory curves of specific models, or choices of $\M\potdep$.
Two examples of these models are shown in \autoref{fig:models}. We see that they have different memory properties.
The serial model performs relatively well at one particular timescale, but it performs poorly at other times.
The cascade model does not perform quite as well at that time, but it maintains its performance over a wider range of timescales.

@@

We now give an upper bound on the initial SNR,
%
\begin{equation}\label{eq:init}
  \snr(0) = \sqrt{N}\prn{2 f\pot f\dep} \eq \prn{\M\pot-\M\dep} \w,
\end{equation}
%
over \emph{all} possible models and also find the class of models that saturate this bound.
A useful quantity is the equilibrium probability flux between two disjoint sets of states, $\CA$ and $\CB$:
%
\begin{equation}\label{eq:flux}
  \F_{\CA\CB} = \sum_{i\in\CA} \sum_{j\in\CB} r \eq_i \frg_{ij}.
\end{equation}
%
The initial SNR is closely related to the flux from the states with $\w_i=-1$ to those with $\w_j=+1$ (see supplementary material):
%
\begin{equation}\label{eq:initflux}
  \snr(0) \leq \frac{4\sqrt{N}\F_{-+}}{r}\,.
\end{equation}
%
This inequality becomes an equality if potentiation never decreases the synaptic weight and depression never increases it, which should be a property of any sensible model.

@@

Now consider the area under the memory curve:
%
\begin{equation}\label{eq:area}
  A = \int_0^\infty\!\!\dr t \, \snr(t).
\end{equation}
%
We will find an upper bound on this quantity as well as the model that saturates this bound.

@@

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \autoref{fig:perts}, see supplementary material for details.
For example, in \autoref{fig:perts}(\ref{fig:tri_right}), for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\smash{\eta^+_i > \eta^+_j}$), we have proven that increasing $\smash{\M\pot_{ij}}$ and decreasing $\smash{\M\dep_{ij}}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \autoref{fig:env}(\ref{fig:sticky}),%\autoref{fig:max_area},
with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some class of perturbations (\eg\ in \autoref{fig:perts}) to further increase the area.

@@

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$ (see supplementary material):
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn[{k - \sum_j j\eq_j}]$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \autoref{fig:env}(\ref{fig:sticky}). %\autoref{fig:max_area}.
This makes it very difficult to exit these states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

@@

This leaves the question of whether this bound is achievable.
At any time, can we find a model whose memory curve touches the envelope?
The red curves in \autoref{fig:env}(\ref{fig:envplot}) show the closest we have come to the envelope with actual models, by repeated numerical optimization of $\SNR(t_0)$ over $\M\potdep$ with random initialization and by hand designed models.

@@

%Another important consideration is the tolerance of the synapses to biological noise, which could be investigated by imposing lower limits on the level of stochasticity of the models.
Finally, it would be interesting to design physiological experiments in order to perform optimal systems identification of potential Markovian dynamical systems hiding within biological synapses,
given measurements of pre and post-synaptic spike trains along with changes in post-synaptic potentials.
Then given our theory, we could match this measured synaptic model to optimal models to understand for which timescales of memory, if any, biological synaptic dynamics may be tuned.
